How predictable is extinction?

### 2018-12-20 doc/intro.tex

By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions. We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa).

By including and modeling this natural variation in extinction risk, we are able to improve our overall predictions because we incorporate and explictly model differences between observations from across a range of possible intensities and selectivies

or the intrisic risk of extinction -- all species can go extinct at any ``moment,'' what are the relative risks between groups and over time
if we gather the kinds of predictors that apparently matter, and use them to make specific predictions of extinction 
If we gather the kinds of predictors that have been found to correlate strongly with extinction risk, how well can we predict species extinction?

Past studies have focused on identifying and measuring the effect of various predictors on extinction risk e.g. does difference in diet or geographic range affect differences in extinction risk?
Here we focus on the quality of our predictions e.g. how accurately can we predict if one species is more likely to go extinct than another?
increasing focus on prediction -- how accurate are we?
understanding versus prediction
long history focused on what determinates are important, but not on how accurate our predictions are given these factors.
the evaluate how well we make predictions in the present
trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!

has been suggested CITE seth wolfgang michael
incorporating extinction survival data from the fossil record can aide in predicting extinction risk of extant species -- the present must at some level be a function of the past
what has not been evaluated is given selectivity changes and intensity changes, how accurate are assessements based on the past likely to be in the future.

for this kind of excercise the best resolved fossil record of skeletonihze marine plantonic microorganisms (names). high temporal resolution, lots of samples, internally consistent taxonomy. 
Rich fossil record -- we might be able to better understand predicting extinction 
trajectories studied before as phenonmenon, but not as tool -- we are using this information as tool!
To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. 
cite the papers that evalutae the importance of the predcitors.
60 million year record lets us ask the 3 main things -- accurate, how much variation in accuracy,

there have been major climate and oceanographic events that have prfoundly influenced the macroevolution of these groups CITATIONS.
intervals of time that are particularly 
things have changed over time -- these moments of big change are ``different'' so predicting them well is porentially important

the amoung of predictive power that the past has at any given 
just the last million years may matter -- so change to range matters
predictions to be better when similar stress over long period of time -- better on the average
a more protracted event/transition (Eo-Mi) -- might expect there to be more predictive power then because past might matter more (key is protraction)

\end{document}

>>>>

By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions. We that that extinction risk varies over time in both intensity (average rate) and selectivity (difference in risk between taxa). Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that varies over time is an important dynamic which shapes the rate and structure of extinction CITATION. What has not been evaluated is as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range of possible intensities and selectivies

By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past. Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk CITATION or on how to identify or measure these effects CITATION. This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are. For example, while a predictor may be considered a ``significant'' factor in comparing the odds of extinction risk between groups, the difference that predictor makes in the probability of extinction can be minimal CITATION STATS TEXTBOOK LOGISTIC REGRESSION. By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.

A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction. For example, we know that a species' global geographic range changes over its duration CITATION. We also know that a species' geographic range size is a good predictor of differences extinction risk CITATION. This begs the question: how does species' extinction risk change over their duration? While the phenomenon of species' geographic range change over time has been studied CITATION FOOTE LIOW WOLFGANG, the potential predictive impact of this change has been under-evaluated CITATION SEE WOLFGANG. For example, does a species' extinction risk increase if they decreased in global geographic range size over 1 million years? Here, we explicitly model and quantify the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction. Similarly, we include species geologic age at time of observation as a potential predictor of extinction -- a factor that may or may not contribute to differences in species extinction risk over time CITATION. Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.

For this kind of exercise, we chose to analyze what is the longest continuous and best resolved fossil record -- that of skeletonized marine plantonic microorganisms from the Cenozoic such as Foraminifera, Radiolarians, Diatoms, and calcareous nannofossils (e.g. coccolithophores). This data is available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project CITATION. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts.

In particular, there have been multiple major climatic and oceanographic events, both short and protracted in duration, over the course of the Cenozoic. For example, the Paleocene-Eocene Thermal Maximum (PETM) was an geologically close-to instantaneous global warming event believed to have been caused by a massive injection of carbon into the atmosphere over a 20,000 year period which rapidly warmed and is associated with substantial changes to the global distribution of species CITATION. In contrast, the Eocene-Oligocene boundary is associated with a protracted global cooling event caused by the slow opening of Drake's passage and the development of the Antarctic ice sheet CITATION. These moments of large scale global change are inherently ``different,'' so predicting extinction before, during, and after these is important -- especially if those events are analogous to future climate change.

Additionally, we would expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions. Alternatively, we would expect our model based predictions of extinction surrounding the PETM may be less accurate because there are inherently fewer samples from the climatic event.

Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

### 2018-12-20 doc/results.tex
full set of cv figures for all datasets

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich ``past and vary'' model has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich ``past and vary'' model is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is then hard to conclude that there is one ``best'' model which we can rely upon. 
% ROC model comparison
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist}
  \caption{Posterior predictive AUC estimates for each of the four models being compared. These estimates are calculated from each of the models posterior predictive distribution compared to the empirical values. Models with a higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1.}
  \label{fig:auc_hist}
\end{figure}

@@

When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria. 
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_taxon}
\end{figure}

Finally, when the posterior predictive distributions of the in-sample AUC estimates are presented over time and by taxonomic group, the reasons for the differences in model adequacy between the taxonomic groups becomes more apparent (Fig. \ref{fig:auc_taxon_time}), in particular the marginally better performance of the ``past and vary'' model over the other three.

@@

@@

@@

%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
%% risk estimate compared to change in geo-range
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
%    \caption{A}
%    \label{fig:relrisk_range}
%  \end{subfigure}
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
%    \caption{B}
%    \label{fig:relrisk_ext}
%  \end{subfigure}
%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation. Geographic range is assumed observed without error. Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
%  \label{fig:relrisk_compare}

>>>>

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich ``past and vary'' model has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich ``past and vary'' model is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is then hard to conclude that there is one ``best'' model which we can rely upon.

% ROC model comparison
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist}
  \caption{Posterior predictive AUC estimates for each of the four models being compared. These estimates are calculated from each of the models posterior predictive distribution compared to the empirical values. Models with a higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1.}
  \label{fig:auc_hist}
\end{figure}

@@

When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria.

We can also compare the by-taxon in-sample performances of the four models.

Finally, we can also examine how the by-taxon in-sample performances of the four models vary over time.

The approximate out-of-sample predictive performance for each of our four models, as measured by AUC, are approximately identical to each other and to our in-sample performance (Fig. \ref{fig:fold_auc}). The quality of performance, however, is not great, with average out-of-sample AUC for each of our models estimated to be about 0.77 which is far from perfect. This result means that while we expect our model to yield consistent results when provided with new data, we do not expect that our predictions will be very accurate.
% ROC OOS estimate

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_taxon}
\end{figure}

@@

We can also compare expected out-of-sample performance based on taxonomic group

Finally, expected out-of-sample performance is compared between taxonomic groups over time.

@@

%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
%
%
%% risk estimate compared to change in geo-range
%\begin{figure}[ht]
%  \centering
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
%    \caption{A}
%    \label{fig:relrisk_range}
%  \end{subfigure}
%
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
%    \caption{B}
%    \label{fig:relrisk_ext}
%  \end{subfigure}
%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation. Geographic range is assumed observed without error. Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
%  \label{fig:relrisk_compare}
%\end{figure}

@@
